---
title: "Investigation into the Exponential Distribution and the Central Limit Theorem"
author: "Sean Clarke"
date: "24 July 2015"
output: pdf_document
---


# Overview

This report is a brief study intended to investigate how the exponential distribution adheres to the **Central Limit Theorem(CLT)**. The CLT states that the distribution of the sum (or mean) of a large number of independent, identically distributed variables will be approximately normal irrespective of the underlying distribution.

# Simulations

We now proceed to perform simulations to investigate how the sample mean and variance approximate the population variance and  mean. We take mutltiple samples of the mean of 40 exponentials.From the central limit theorem, we would expect:


$$E[\overline{X}] = \mu_{Pop}$$ 

$$Var[\overline{X}] = \sigma^{2}_{Pop}/n$$

The Exponential distribution has PDF

$$  f(x, \lambda)=\begin{cases}
    \lambda e^{-\lambda x}, & \text{if $x>=0$}.\\
    0, & \text{if $x<0$}.
  \end{cases}$$
  
where

$$ E[X] = 1/\lambda $$
$$ Var[X] = 1/\lambda^{2} $$

In this analysis we calculate the exponentials using the R function $rexp(n, \lambda)$ where we fix $\lambda = 0.2$. Firstly we generate the data set.

```{r}
numberSims <- 1000
numberMeans <- 40
lambda <- 0.2
mns <- NULL
for (i in 1 : numberSims) {mns <- c(mns, mean(rexp(numberMeans,lambda)))}
```
## Sample Mean versus Theoretical Mean 

Initially we consider comparison between the mean calculated from the data set and what the mean should be according to the CLT. So according to the CLT, for a sufficiently large data set, the sample means should follow a normal distribution with mean $\mu$, for the exponential distribution:

$$ E[\overline{X}] = \mu = 1/\lambda$$

Then we proceed to calculate the mean of dataset

```{r}
c(1/lambda,mean(mns))
```

So we the mean calculated in the simulation is very close to the theoretical mean.

## Sample Variance versus Theoretical Variance

According to the CLT, the variance for means of an Exponential distribution is given by

$$ V[\overline{X}] = \sigma^{2}/n = (1/\lambda)^{2}/n = 1/\lambda^{2}n = $$

```{r}
c(1/(lambda^2*numberMeans), var(mns))
```

Again we see that the theoretical mean is very close to that obtained from the simulation.

#Distribution: Via figures and text, explain how one can tell the distribution is approximately normal.